---
title: "SQLDataFrame: Internal Implementation"
author:
- name: Qian Liu
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "last compiled: `r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
package: SQLDataFrame
vignette: >
  %\VignetteIndexEntry{SQLDataFrame Internal Implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
date: "last edit: 4/3/2018"

[dbplyr]: https://cran.r-project.org/web/packages/dbplyr/index.html


# SQLDataFrame slots

```{r constructor}
dbfile <- system.file("extdata/test.db", package = "SQLDataFrame")
obj <- SQLDataFrame(
    dbname = dbfile, dbtable = "state", dbkey = "state")
obj
```

## Slot accessors

To make the `SQLDataFrame` object as light and compact as possible,
there are only 5 slots contained in the object: `tblData`, `dbkey`,
`dbnrows`, `dbconcatKey`, `indexes`. Metadata information could be
returned through these 5 slots using slot accessors or other utility
functions. 
```{r}
slotNames(obj)
dbname(obj)
dbtable(obj)
dbkey(obj)
dbconcatKey(obj)
```

## `tblData` slot

The `tblData` slot saves the `dbplyr::tbl_dbi` version of the database
table, which is a light-weight representation of the database table in
_R_. Of note is that this lazy tbl only contains unique rows. It could
also be sorted by the `dbkey(obj)` if the `SQLDataFrame` object was
generated from `union` or `rbind`. So when the `saveSQLDataFrame()`
function was called, a database table will be written into a physical
disk space and have the unique records.

```{r}
obj@tblData
```

## `dbnrows` and `dbconcatKey`
The `dbnrows` slot saves the number of rows corresponding to the
`tblData`, and `dbconcatKey` saves the realized (concatenated if
multiple) key columns corresponding to the `tblData`. 

```{r}
obj@dbnrows
obj@dbconcatKey
```

## `indexes` slot
The `indexes` slots is an unnamed list saving the row and column
indexes respectively corresponding to the `tblData` slot, so that the
`SQLDataFrame` could possibly have duplicate rows or only a subset of
data records from the `tblData`, while the `tblData` slot doesn't need
to be changed. To be consistent, the slots of `dbnrows` and
`dbconcatKey` will also remain unchanged.

```{r}
obj@indexes
obj_sub <- obj[sample(5, 3, replace = TRUE), 2:3]
obj_sub
obj_sub@indexes
identical(obj@tblData, obj_sub@tblData)
```

With a `filter` function (which is similar to `[i, ]` subsetting),
only the `indexes` slot will be updated for the row index pointing to
the `tblData`.

```{r}
obj_filter <- obj %>% filter(division == "South Atlantic" & size == "medium")
obj_filter@indexes
identical(obj@tblData, obj_filter@tblData)
```

# SQLDataFrame methods

## ROWNAMES

The `ROWNAMES,SQLDataFrame` method was defined to return the
(concatenated if multiple) key column(s) value, so that the row
subsetting with character vector works for the `SQLDataFrame` objects.

```{r}
rnms <- ROWNAMES(obj)
obj[sample(rnms, 3), ]
```

For `SQLDataFrame` object with composite keys: 

```{r}
obj1 <- SQLDataFrame(dbname = test.db, dbtable = "state",
                     dbkey = c("region", "population"))
ROWNAMES(obj1[1:10,])
obj1[c("South\b3615.0", "West\b365.0"), ]
```

# SessionInfo()

```{r}
sessionInfo()
```
