---
title: "SQLDataFrame: Representation of database table in DataFrame metaphor"
author:
- name: Qian Liu
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "last compiled: `r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
package: SQLDataFrame
vignette: >
  %\VignetteIndexEntry{SQLDataFrame: Representation of database table in DataFrame metaphor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
date: "last edit: 11/30/2018"

# Introduction


Theoretically, users could choose to use different database backend
when doing the `SQLDataFrame` constructor. Five commonly used
backends are:

- _RMySQL_: connects to MySQL and MariaDB
- _RPostgreSQL_: connects to Postgres and Redshift.
- **RSQLite**: embeds a SQLite database. (now used as default)
- _odbc_ connects to many commercial databases via the open database
  connectivity protocol.
- _bigrquery_ connects to Google’s BigQuery.

[dbplyr]: https://cran.r-project.org/web/packages/dbplyr/index.html

# Installation

1. Download the package. 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SQLDataFrame")
```
The development version is also available to download from Github. 
```{r getDevel, eval=FALSE}
BiocManager::install("Liubuntu/SQLDataFrame")
```

2. Load the package into R session.
```{r Load, message=FALSE}
library(SQLDataFrame)
```

# SQLDataFrame
## SQLDataFrame constructor
To construct a `SQLDataFrame` object, 3 arguments are needed:
`dbname`, `dbtable` and `dbkey`. The `dbname` is the file path to the
database that is saved on-disk. `dbtable` argument specifies the
database table name that is going to be represented in `SQLDataFrame`
object. If only one table is available in the specified database name,
this argument could be left blank. The `dbkey` argument is used to
specify the column name in the table which could uniquely identify all
the data observations (rows).  
The `show` method for `SQLDataFrame` has consistent format with the 
`DataFrame` object.


```{r constructor}
test.db <- system.file("inst/extdata/test.db", package = "SQLDataFrame")
obj <- SQLDataFrame(
    dbname = test.db, dbtable = "colDatal", dbkey = "sampleID")
obj
```

## SQLDataFrame methods
### Slot accessors

To make the `SQLDataFrame` object as light and compact as possible,
there are only 6 slots contained in the object. We can obtain specific
slot with the corresponding slot accessor functions.
```{r}
slotNames(obj)
dbtable(obj)
dbkey(obj)
colnames(obj)
nrow(obj)
```
The `tblData` slot saves the `dbplyr::tbl_dbi` version of the database
table, which is a light-weight representation of the database table in
_R_. The `dbrownames` and `dbnrow` slot saves the row information for
the original database table. When subsetting `SQLDataFrame` objects
using `[`, these 3 slots remains unchanged, but the `nrow()`,
`rownames()` and `show()` method will reflect the data subsetting.

```{r}
obj@tblData
obj@dbrownames
obj@dbnrows
```

The `dbname()` method returns the absolute path for the database. It
was extracted from the `obj@tblData` instead of being a stand-alone
slot.

```{r}
dbname(obj)
```
Besides, many more useful methods are defined to get general
information from `SQLDataFrame` object. e.g., we can use `dimnames()`
to return the row/colnames of the data. It returns an unnamed list,
with the first element being rownames (could also use `rownames()`
method), and 2nd element being colnames (could also use `colnames()`
method). Could also use `rownames()/colnames()` to extract the
specific dimension names. `dim()` returns the dimension of the
database table. Could also use `nrow()/ncol()` to extract a specific
dimension. `length()` method is also defined which works same as
`ncol()`.   
By default, the `rownames(SQLDataFrame)` would be `NULL` as row names
are normally undesirable when data is large. But people could add it
using the constructor`SQLDataFrame(row.names=)`. 

```{r methods}
dim(obj)
identical(dim(obj), c(nrow(obj), ncol(obj)))
dimnames(obj)
length(obj)
```

### `[[` subsetting
Behaves similarly to `[[,DataFrame` and returns a realized vector of
values from a single column. `$,SQLDataFrame` is also defined to
conveniently extract column values. 

```{r}
obj[[1]]
obj[["ages"]]
identical(obj[[3]], obj[["ages"]])
obj$ages
```

### `[` subsetting
`SQLDataFrame` instances can be subsetted in a similar way of
`DataFrame` following the usual _R_ conventions, with numeric,
character or logical vectors; logical vectors are recycled to the
appropriate length.

```{r, subsetting}
obj[1:3, 1:3]
obj[c(TRUE, FALSE), c(TRUE, FALSE)]
obj[1:3, c("sampleID", "ages")]

obj1 <- obj[1:3, 1:3]
```

List style subsetting is also allowed to extract certain columns from
the `SQLDataFrame` object which returns `SQLDataFrame` by default.  
```{r}
obj[1]
obj[c("sampleID", "ages")]
```
When subsetting using `[` for a single column, the default is to
return the realized value, unless user specify `drop=FALSE` explicitly. 
```{r}
obj[, 1]
obj[, 1, drop=FALSE]
```

### Numeric calculation / Data manipulation (todo)

Numeric calculations could be evaluated on `SQLDataFrame` objects.

```{r, numeric, eval=FALSE}
sum(obj@ages)
```

# SessionInfo()

```{r}
sessionInfo()
```


