---
title: "SQLDataFrame: Lazy representation of database table in DataFrame metaphor"
author:
- name: Qian Liu
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "last compiled: `r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
package: SQLDataFrame
vignette: >
  %\VignetteIndexEntry{SQLDataFrame: Lazy representation of database table in DataFrame metaphor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
date: "last edit: 4/5/2018"

# Introduction

As the genomic data are getting increasingly large. With the memory
limit of _R_, it's getting more and more difficult to manipuate large
data. the `matrix`-like high-throughput genomic data are now being able to easily loaded and
manipulated within R by extending[DelayedArray][] to represent on-disk
data structures with different back-end in R-user-friendly array data
structure (e.g.,[HDF5Array][], [GDSArray][]). However, the annotation
file usually have`data.frame`-like structure, and include meta info with different data type. Here we
 have developed an _Bioconductor_ package`SQLDataFrame` to have the `DataFrame` metaphor with database table backend.

Currently we are only supporting the DBI backend of `RSQLite`. But
theoretically, we would implement this package so that users could
choose to use different database backend when doing the `SQLDataFrame`
constructor and `saveSQLDataFrame`. 

Support of other backends will be added for next version. Five
commonly used backends are:

- **RSQLite**: embeds a SQLite database. (now used as default)
- _RMySQL_: connects to MySQL and MariaDB
- _RPostgreSQL_: connects to Postgres and Redshift.
- _odbc_ connects to many commercial databases via the open database
  connectivity protocol.
- _bigrquery_ connects to Google’s BigQuery.

[dbplyr]: https://cran.r-project.org/web/packages/dbplyr/index.html
[DelayedArray]: https://bioconductor.org/packages/DelayedArray
[GDSArray]: https://bioconductor.org/packages/GDSArray
[HDF5Array]: https://bioconductor.org/packages/HDF5Array

# Installation

1. Download the package. 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SQLDataFrame")
```
The development version is also available to download from Github. 
```{r getDevel, eval=FALSE}
BiocManager::install("Liubuntu/SQLDataFrame")
```
2. Load the package(s) into R session.
```{r Load, message=FALSE, eval = TRUE}
library(SQLDataFrame)
library(DBI)
```

# `SQLDataFrame` class

## `SQLDataFrame` constructor
To construct a `SQLDataFrame` object, 3 arguments are needed:
`dbname`, `dbtable` and `dbkey`. The `dbname` is the file path to the
database that is on-disk or remote. `dbtable` argument specifies the
database table name that is going to be represented as `SQLDataFrame`
object. If only one table is available in the specified database name,
this argument could be left blank. The `dbkey` argument is used to
specify the column name in the table which could uniquely identify all
the data observations (rows).  

Note that after reading the database table into `SQLDataFrame`, the
key columns will be kept as fixed columns showing on the left hand
side, with `|` separating key column(s) with the other columns. The
`ncol`, `colnames`, and corresponding column subsetting will only
apply to the non-key-columns.


```{r constructor}
dbfile <- system.file("extdata/test.db", package = "SQLDataFrame")
conn <- DBI::dbConnect(dbDriver("SQLite"), dbname = dbfile)
obj <- SQLDataFrame(
    conn = conn, dbtable = "colData", dbkey = "sampleID")
obj
dim(obj)
colnames(obj)
```

## Slot & accessors

To make the `SQLDataFrame` object as light and compact as possible,
there are only 5 slots contained in the object: `tblData`, `dbkey`,
`dbnrows`, `dbconcatKey`, `indexes`. Metadata information could be
returned through these 5 slots using slot accessors or other utility
functions.  

```{r}
slotNames(obj)
dbtable(obj)
dbkey(obj)
connSQLDataFrame(obj)
```

Besides, many useful common methods are defined on `SQLDataFrame`
object to make it a more DataFrame-like data structure. e.g., we can
use `dimnames()` to return the row/colnames of the data. It returns an
unnamed list, with the first element being rownames which is always
`NULL`, and 2nd element being colnames (could also use `colnames()`
method). `dim()` method is defined to return the dimension of the
database table, which enables the `nrow()/ncol()` to extract a
specific dimension. `length()` method is also defined which works same
as `ncol()`.  
Note that the `rownames(SQLDataFrame)` would always be `NULL` as
rownames are not supported in `SQLDataFrame`. However, `ROWNAMES(obj)`
was implemented for the `[` subsetting with characters. 
 
```{r methods}
dim(obj)
dimnames(obj)
length(obj)
ROWNAMES(obj)
```

**NOTE** that the `dbtable()` accessor only works for a `SQLDataFrame`
object that the lazy tbl carried in `tblData` slot corresponds to a
single database. If the `SQLDataFrame` was generated from `rbind`,
`union` or `*_join`, call `saveSQLDataFrame()` to save the lazy tbl to
disk so that `dbtable()` will be activated.

```{r}
dbtable(obj)
aa <- rbind(obj[1:5, ], obj[6:10, ])
aa
dbtable(aa)  ## message
bb <- saveSQLDataFrame(aa, dbname = tempfile(fileext=".db"),
                       dbtable = "aa", overwrite = TRUE)
connSQLDataFrame(bb)
dbtable(bb)
```

# makeSQLDataFrame

We could also construct a `SQLDataFrame` object directly from a file
name. The `makeSQLDataFrame` function takes input of character value
of file name for common text files (.csv, .txt, etc.), write into
database tables, and open as `SQLDataFrame` object. Users could
provide values for the `dbname` and `dbtable` argument. If NULL,
default value for `dbname` would be a temporary database file, and
`dbtable` would be the `basename(filename)` without extension.

**NOTE** that the input file must have one or multiple columns that
could uniquely identify each observation (row) to be used the
`dbkey()` for `SQLDataFrame`. Also the file must be rectangular, i.e.,
rownames are not accepted. But users could save rownames as a separate
column. 

```{r}
mtc <- tibble::rownames_to_column(mtcars)[,1:6]
filename <- file.path(tempdir(), "mtc.csv")
write.csv(mtc, file= filename, row.names = FALSE)
aa <- makeSQLDataFrame(filename, dbkey = "rowname", sep = ",",
                       overwrite = TRUE)
aa
connSQLDataFrame(aa)
dbtable(aa)
```

# saveSQLDataFrame

With all the methods (`[` subsetting, `rbind`, `*_join`, etc.,)
provided in the next section, the `SQLDataFrame` always work like a
lazy representation until users explicitly call the `saveSQLDataFrame`
function for realization. `saveSQLDataFrame` write the lazy tbl
carried in `tblData` slot into an on-disk database table, and re-open
the `SQLDataFrame` object from the new path.

It's also recommended that users call `saveSQLDataFrame` frequently to
avoid too many lazy layers which slows down the data processing. 

```{r}
dbfile <- system.file("extdata/test.db", package = "SQLDataFrame")
conn <- DBI::dbConnect(DBI::dbDriver("SQLite"), dbname = dbfile)
obj <- SQLDataFrame(conn = conn, dbtable = "state", dbkey = "state")
conn
dbtable(obj)
obj1 <- saveSQLDataFrame(obj, dbname = tempfile(fileext = ".db"),
                        dbtable = "obj_copy")
connSQLDataFrame(obj1)
dbtable(obj1)
```

# SQLDataFrame methods

## `[[` subsetting
`[[,SQLDataFrame` Behaves similarly to `[[,DataFrame` and returns a
realized vector of values from a single column. `$,SQLDataFrame` is
also defined to conveniently extract column values.

```{r}
head(obj[[1]])
head(obj[["region"]])
head(obj$size)
```

We can also get the key column values using character extraction. 
```{r}
head(obj[["state"]])
```

## `[` subsetting

`SQLDataFrame` instances can be subsetted in a similar way of
`DataFrame` following the usual _R_ conventions, with numeric,
character or logical vectors; logical vectors are recycled to the
appropriate length. 

**NOTE**, use `drop=FALSE` explicitly for single column subsetting if
you want to return a `SQLDataFrame` object, otherwise, the default
`drop=TRUE` would always return a realized value for that column.

```{r, subsetting}
obj[1:3, 1:2]
obj[c(TRUE, FALSE), c(TRUE, FALSE), drop=FALSE]
obj[1:3, "population", drop=FALSE]
obj[, "population"]  ## realized column value
```

Subsetting with character vector works for the `SQLDataFrame`
objects. With composite keys, users need to concatenate the key values
by `\b` for row subsetting (See the vignette for internal
implementation for more details).

```{r}
rnms <- ROWNAMES(obj)
obj[c("Alabama", "Colorado"), ]
```

```{r}
obj1 <- SQLDataFrame(conn = conn, dbtable = "state",
                     dbkey = c("region", "population"))
rnms <- ROWNAMES(obj1)
obj1[c("South\b3615.0", "West\b365.0"), ]
```

List style subsetting is also allowed to extract certain columns from
the `SQLDataFrame` object which returns `SQLDataFrame` by default.  

```{r}
obj[1]
obj["region"]
```

## filter & mutate

We have also enabled the S3 methods of `filter` and `mutate` from
`dplyr` package, so that users could have the convenience in filtering
data observations and adding new columns.

```{r}
obj1 %>% filter(division == "South Atlantic" & size == "medium")
```

```{r}
obj1 %>% mutate(p1 = population/10, s1 = size)
```


## union & rbind

To be consistent with `DataFrame`, `union` and `rbind` methods were
implemented for `SQLDataFrame`, where `union` returns the
`SQLDataFrame` sorted by the `dbkey(obj)`, and `rbind` keeps the
original orders of input objects.

```{r}
dbfile1 <- system.file("extdata/test.db", package = "SQLDataFrame")
con1 <- DBI::dbConnect(dbDriver("SQLite"), dbname = dbfile1)
dbfile2 <- system.file("extdata/test1.db", package = "SQLDataFrame")
con2 <- DBI::dbConnect(dbDriver("SQLite"), dbname = dbfile2)
ss1 <- SQLDataFrame(conn = con1, dbtable = "state",
                    dbkey = c("state"))
ss2 <- SQLDataFrame(conn = con2, dbtable = "state1",
                    dbkey = c("state"))
ss11 <- ss1[sample(5), ]
ss21 <- ss2[sample(10, 5), ]
```

- union
```{r, eval=FALSE}
obj1 <- union(ss11, ss21) 
obj1  ## reordered by the "dbkey()"
```

- rbind
```{r}
obj2 <- rbind(ss11, ss21) 
obj2  ## keeping the original order by updating the row index
```

**maybe remove this part for vignette.** 
```{r, eval=FALSE}
identical(as.data.frame(tblData(ss11)), as.data.frame(tblData(ss21)))
```

## *_join methods

The `*_join` family methods was implemented for `SQLDataFrame`
objects, including the `left_join`, `inner_join`, `semi_join` and
`anti_join`, which provides the capability of merging database files
from different sources.

```{r}
ss12 <- ss1[1:10, 1:2]
ss22 <- ss2[6:15, 3:4]
left_join(ss12, ss22)
inner_join(ss12, ss22)
semi_join(ss12, ss22)
anti_join(ss12, ss22)
```

# SessionInfo()

```{r}
sessionInfo()
```
