---
title: Using SQL-backed DataFrames
author:
- name: Qian Liu
  email: Qian.Liu@RoswellPark.org
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Aaron Lun
  affiliation: Genentech, Inc.
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
package: SQLDataFrame
date: "Revised: February 5, 2024"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
library(BiocStyle)
self <- Biocpkg("SQLDataFrame");
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Overview

Firstly, I would like to extend my special thanks to Aaron Lun for his
foundational work on the `ParquetDataFrame` package, and for his
highly technical and detailed suggestions aimed at enhancing the
current implementation of the [`SQLDataFrame`
package](https://github.com/Bioconductor/SQLDataFrame/issues/8). Here
I'm introducing the new version of `SQLDataFrame` for handling
different SQL-backed files. Essentially, the implementation of
`SQLDataFrame` is modeled upon `ParquetDataFrame` regarding its data
structure, construction, documentation, and examples. This approach
ensures the retension of best practices and maintains consistentcy in
the use within Bioconductor ecosystem, thus simplifying the learning
curve for users.

The `SQLDataFrame`, as its name suggests, is a `DataFrame` where the
columns are derived from data in a SQL table. This is fully
file-backed so no data is actually loaded into memory until requested,
allowing users to represent large datasets in limited memory. As the
`SQLDataFrame` inherits from `r Biocpkg("S4Vectors")`' `DataFrame`, it
can be used anywhere in Bioconductor's ecosystem that accepts a
`DataFrame`, e.g., as the column data of a `SummarizedExperiment`, or
inside a `BumpyDataFrameMatrix` from the `r Biocpkg("BumpyMatrix")`
package.

`SQLDataFrame` currently supports the following backends with their
respective extension classes (and constructor functions):

- `SQLite`: `SQLiteDataFrame`
- `DuckDB`: `DuckDBDataFrame`
- `Parquet`: `ParquetDataFrame`

Note that the support of `Parquet` is different from the
`ParquetDataFrame` by Aaron. It uses the raw SQL queries for data
extraction instead of through the `arrow` `Dataset`.

It can be easily extended to any other SQL-backed file types by simply
defining the extension classs in `SQL_extensions.R` with minor updates
in `acquireConn.R` to create a database instance, `SQLDataFrame.R`,
and `SQLColumnSeed.R`. Pull requests for adding new SQL backends are
welcome!

# Package installation

1. Download the package from Bioconductor.

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SQLDataFrame")
```

2. Load the package into R session.
```{r Load, message=FALSE}
library(SQLDataFrame)
```

# Basic usage

Given a path, database type and table name to a SQL file, we can
construct a `SQLDataFrame`. If the backend is supported, we can
construct an extension class directly.

```{r}
## Mocking up a file.
tf <- tempfile()
con <- DBI::dbConnect(RSQLite::SQLite(), tf)
DBI::dbWriteTable(con, "mtcars", mtcars)
DBI::dbDisconnect(con)


## Creating the SQLiteDataFrame.
library(SQLDataFrame)
df <- SQLDataFrame(tf, dbtype = "sqlite", table = "mtcars")
df0 <- SQLiteDataFrame(tf, table = "mtcars")
identical(df, df0)
```

Similarly, we can create a `DuckDbDataFrame`: 

```{r}
tf1 <- tempfile()
on.exit(unlist(tf1))
con <- DBI::dbConnect(duckdb::duckdb(), tf1)
DBI::dbWriteTable(con, "mtcars", mtcars)
DBI::dbDisconnect(con)

df1 <- SQLDataFrame(tf1, dbtype = "duckdb", table = "mtcars")
df2 <- DuckDBDataFrame(tf1, table = "mtcars")
identical(df1, df2)
```

We can also create a `ParquetDataFrame`. NOte that the `table`
argument is not required for the constructor. Any given value will be
overwritten by `my_parquet_table`. 

```{r}
tf2 <- tempfile()
on.exit(unlink(tf2))
arrow::write_dataset(mtcars, tf2, format = "parquet")
dir(tf2)
 
df3 <- SQLDataFrame(tf2, dbtype = "parquet") ## does not request "table"
df4 <- ParquetDataFrame(tf2)  ## does not request "table"
identical(df3, df4)
```

These support all the usual methods for a `DataFrame`, except that the
data is kept on file and referenced as needed:

```{r}
nrow(df)
colnames(df)
class(as.data.frame(df))
```

We extract individual columns as `SQLColumnVector` objects. These
are 1-dimensional file-backed `DelayedArray`s that pull a single
column's data from the SQL table on demand.

```{r}
df$mpg

# These can participate in usual vector operations:
df$mpg * 10
log1p(df$mpg)

# Realize this into an ordinary vector.
as.vector(df$mpg)
```

# Collapsing to a `DFrame`

The main goal of a `SQLDataFrame` is to serve as a consistent
representation of the data inside a SQL table. However, this cannot be
easily reconciled with many `DataFrame` operations that add or change
data - at least, not without mutating the SQL file, which is outside
the scope of the `SQLDataFrame` class. To handle such operations,
the `SQLDataFrame` will collapse to a `DFrame` of
`SQLColumnVector`s:

```{r}
copy <- df
copy$some_random_thing <- runif(nrow(df))
class(copy)
colnames(copy)
```

This preserves the memory efficiency of file-backed data while
supporting all `DataFrame` operations. For example, we can easily
subset and mutate the various columns, which manifest as delayed
operations inside each column.

```{r}
copy$wt <- copy$wt * 1000
top.hits <- head(copy)
top.hits
```

The fallback to `DFrame` ensures that a `SQLDataFrame` is
interoperable with other Bioconductor data structures that need to
perform arbitrary `DataFrame` operations. Of course, when a collapse
occurs, we lose all guarantees that the in-memory representation is
compatible with the underlying SQL table. This may preclude further
optimizations in cases where we consider directly operating on the
file.

# Retrieving the SQL connection

At any point, users can retrieve a handle of connection to the
underlying SQL file via the `acquireConn()` function. This can be used
with methods in the `r CRANpkg("DBI")`, `r CRANpkg("RSQLite")`, and `r
CRANpkg("duckdb")` packages. The cached `DBIConnection` for any given
`path` can be deleted by calling `releaseConn`.

Note that for Parquet table, it is opened as an `arrow Dataset` first
and then registered in a duckdb connection as a virtual
table. `releaseConn` will unregister the table and close the
connection.

```{r}
handle <- acquireConn(path(df), dbtype = dbtype(df))
handle
releaseConn(path(df))
```

Note that the acquired handle will not capture any delayed
subsetting/mutation operations that have been applied in the R
session. In theory, it is possible to convert a subset of `r
Biocpkg("DelayedArray")` operations into their `r CRANpkg("DBI")`
equivalents, which would improve performance by avoiding the R
interpreter when executing a query on the file.

In practice, any performance boost tends to be rather fragile as only
a subset of operations are supported, meaning that it is easy to
silently fall back to R-based evaluation when an unsupported operation
is executed. Users wanting to optimize query performance should just
operate on the handle directly.

# Session information {-}

```{r}
sessionInfo()
```
